A decomposition Approach for solving the JSP using multi-shot ASP\thanks{Supported by organization x.}

### 2022-01-30 ICLP Paper/samplepaper.tex
Section 4

\newcommand{\clingo}{\emph{clingo}\xspace}
\newcommand{\clingodl}{\emph{clingo}[DL]\xspace}
\newcommand{\figspace}{\vspace{-1ex}}

\makeatletter
\def\fcapsize@figure{\normalfont\normalsize\rmfamily}
\def\fcapstyle@figure{\normalfont\normalsize\itshape}
\makeatother

@@

The \lstinline{step(w)} subprogram until line~\ref{prg:diff_log:oper_limit:end} 
constitutes the central part of our multi-shot ASP modulo DL encoding in Listing~\ref{prg:encoding}, whose parameter~\lstinline{w} stands for consecutive integers from~\lstinline{1}
identifying time windows to schedule.
Auxiliary atoms of the form \lstinline{use(}$m$\lstinline{,}$w'$\lstinline{,}$w$\lstinline{)},
supplied by the rules in lines~\ref{prg:encoding:use1} and~\ref{prg:encoding:use2},
indicate the latest time window $1\leq w'\leq w$ including some operation that needs
to be processed by machine~$m$.
The next rule in lines \ref{prg:base:sharing_machine:begin}-\ref{prg:base:sharing_machine:end}
identifies pairs \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} and
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)} of operations sharing
the same machine~$m$,
where
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)} belongs to the time window~$w$ and
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} is either 
(i)
contained in the latest time window $1\leq w'< w$ indicated by
\lstinline{use(}$m$\lstinline{,}$w'$\lstinline{,}$w-1$\lstinline{)} or
(ii)
also part of the time window~$w$, in which case $j_1 < j_2$ establishes
an asymmetric representation for the pair of operations in derived atoms
\lstinline{share((}$j_1$\lstinline{,}$s_1$\lstinline{),(}$j_2$\lstinline{,}$s_2$\lstinline{),}$p_1$\lstinline{,}$p_2$\lstinline{,}$x$\lstinline{,}$w$\lstinline{)}.
If the flag $x=\text{\lstinline{1}}$ signals that 
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)}
belongs to a previous time window~$w'$,
the rule in line~\ref{prg:base:ordered} % applies and 
derives the atom
\lstinline{order((}$j_1$\lstinline{,}$s_1$\lstinline{),(}$j_2$\lstinline{,}$s_2$\lstinline{),}$p_1$\lstinline{,}$w$\lstinline{)}
to signal that \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} needs to be completed
before performing \lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)}, i.e.,
the execution order must comply with the decomposition into time windows.
The rule in lines \ref{prg:base:same_job:begin}-\ref{prg:base:same_job:end} yields
a similar atom when \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)}
is the predecessor operation $s_1 = s_2-1$ of \lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)}
in the same job $j_1=j_2$.
In contrast to the cases in which 
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} must be processed before
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)},
the choice rule in line~\ref{prg:base:order}
allows for performing two operations sharing a machine in the lexicographic order
of their jobs if the operations belong to the same time window.
In case the atom representing execution in lexicographic order is not chosen,
the rule in lines \ref{prg:base:order:begin}-\ref{prg:base:order:end}
derives an atom signaling the inverse, given that the operations must not intersect and some sequence has to be determined.%
%
\lstinputlisting[float=t,label=prg:encoding,caption={Multi-shot ASP modulo DL encoding}]{listing/encoding.lp}

@@

\begin{table}[t]
    \caption{%
      Experiments varying the number of time windows on JSP benchmark sets of three sizes\label{tab:Table01}}%
      % Results for three sizes of job shop scheduling instances with different numbers of time windows using static machine-based EST decomposition \label{tab:Table01}}%
%    \mbox{~}\hfill
%    \resizebox{0.8\textwidth}{!}{%
%    \begin{minipage}{\textwidth}
    \centering
    \begin{tabular}{r l r r r r r r r}
    \topline%\hline
%       & {}        & \multicolumn{6}{c}{Group of Instances}  \\
%    &  &  \multicolumn{7}{c}{Number of Time Windows} \\
    Size & Average & 1 & 2 & 3 & 4 & 5 & 6 & 10 % \\

@@

    {$50\times15$}       & Makespan             & $3542.1$   & $3149.7$	     &  \boldmath{$3083.8$} & $3110.7$ & $3213.6$ & $3225.3$ & $3524.9$ \\
    & Time             & $1000.0$	& $1000.0$   & $1000.0$	 & $934.4$ & $615.5$  & $243.8$  & $18.5$  \\
    & Interrupts    & $1.0$	    & $2.0$      & $3.0$	& $3.6$  & $2.7$  & $1.0$  & $0.1$       \\[1.5mm]

@@

    % \\[-2.75mm]
    \botline%\hline
    \end{tabular}
   % }
%    \end{minipage}
% }
%    \hfill\hfill\mbox{~}
\end{table}
%
Our first series of experiments shown in Table~\ref{tab:Table01} concerns the number
of time windows into which the $750$, $1000$, or $2000$ operations, respectively,
of the instances with $50\times15$, $50\times20$, and $100\times20$ jobs and machines
are split.
We use the static Machine-based EST decomposition strategy, which augments the encoding
given in Listing~\ref{prg:decomposition} with bottleneck machines for inserting 
their operations into time windows, as baseline because it leads to more robust
solution quality than Job-based problem decomposition.
For each of the three problem sizes,
we report the average makespan over ten instances,
where smaller values indicate better schedules,
the average runtime of \clingodl, and the
average number of interrupted optimization runs on subproblems.
We gradually increase the number of time windows from~$1$ to~$6$ and additionally
include results for~$10$ time windows to outline the trend of degrading
solution quality when the partition into time windows becomes too fragmented
for merging (optimized) partial schedules into proficient global solutions.
In fact, the shortest makespans, highlighted in boldface,
are obtained with problem decomposition into
$3$, $4$, or $6$ time windows, respectively, for the instances of size 
$50\times15$, $50\times20$, and $100\times20$.
Compared to the results for $1$ time window,
which represent global optimization on the full problem up to the
runtime limit of $1000$ seconds,
we observe substantial makespan improvements, amounting to almost one order
of magnitude for the largest instances with $100\times20$ jobs and machines.
These advantages do not come as a surprise considering that the JSP instances
are highly combinatorial \citep{shysha18a} and each global optimization run
times out with a more or less optimized solution, 
where a good-quality schedule is the more difficult to find the greater the
number of operations gets.
Except for about 12\% completed optimization runs in multi-shot solving with
$4$ time windows on the $50\times20$ instances,
the best schedules are obtained with optimization runs timing out on subproblems,
yet with good-quality partial schedules that can be merged into better global
solutions than achievable with more time windows, which
become smaller and easier to solve but also cut off good-quality
schedules.

@@

\begin{table}[t]
    \caption{Experiments comparing Job- and Machine-based problem decomposition strategies\label{tab:Table02}}%
%    \mbox{~}\hfill
%    \resizebox{0.8\textwidth}{!}{%
%    \begin{minipage}{\textwidth}
    \centering
    \begin{tabular}{r r r r r r r r r r}
    \topline% \hline

@@

    % \\[-2.75mm]
    \botline% \hline
    \end{tabular}
   % }
%    \end{minipage}
% }
%    \hfill\hfill\mbox{~}
\end{table}
%
In Table~\ref{tab:Table02}, we compare other problem decomposition strategies
to the static Machine-based EST approach, providing the baseline abbreviated by M-EST.
Further decomposition strategies are denoted by M-MTWR for the static Machine-based MTWR
variant, corresponding dynamic versions M-EST$^\text{\,d}$ and M-MTWR$^\text{\,d}$ that
determine EST and MTWR values w.r.t.\ partial schedules from previous time windows, as well as
static Job-based counterparts J-EST and J-MTWR in the first two rows.
For each decomposition strategy and JSP instance size, we present results for the number
of time windows,
which amounts to the next greater integer of the average number of interrupted optimization
runs in columns headed by Ints, leading to the shortest average makespan.
The considerably increased average makespans in the first two rows clearly indicate that
time windows determined with Job-based decomposition strategies are less adequate than
those investigating bottleneck machines in the first place, and then picking their operations
based on the smallest EST or greatest MTWR value as a secondary criterion.
While Machine-based decomposition turns out to be advantageous,
there is no clear gap between the static M-EST and M-MTWR variants as well as their dynamic
versions M-EST$^\text{\,d}$ and M-MTWR$^\text{\,d}$.
That is, the heuristic to schedule operations from highly loaded machines
as soon as possible works well for decomposing the considered JSP instances,
yet further solution quality differences due to minor deviations in the
decomposition strategy or optimization performance of \clingodl are negligible.

>>>>

\newcommand{\clingo}{\emph{clingo}\xspace}
\newcommand{\clingodl}{\emph{clingo}[DL]\xspace}
\newcommand{\figspace}{\vspace{-1ex}}
\newcommand{\ksverify}[1]{\textcolor{red}{#1}}
\makeatletter
\def\fcapsize@figure{\normalfont\normalsize\rmfamily}
\def\fcapstyle@figure{\normalfont\normalsize\itshape}
\makeatother

@@

The \lstinline{step(w)} subprogram until line~\ref{prg:diff_log:oper_limit:end} 
in Listing~\ref{prg:encoding} constitutes the central part of our multi-shot ASP modulo DL encoding, whose parameter~\lstinline{w} stands for consecutive integers from~\lstinline{1}
identifying time windows to schedule.
Auxiliary atoms of the form \lstinline{use(}$m$\lstinline{,}$w'$\lstinline{,}$w$\lstinline{)},
supplied by the rules in lines~\ref{prg:encoding:use1} and~\ref{prg:encoding:use2},
indicate the latest time window $1\leq w'\leq w$ including some operation that needs
to be processed by machine~$m$.
The next rule in lines \ref{prg:base:sharing_machine:begin}-\ref{prg:base:sharing_machine:end}
identifies pairs \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} and
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)} of operations sharing
the same machine~$m$,
where
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)} belongs to the time window~$w$ and
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} is either 
(i)
contained in the latest time window $1\leq w'< w$ indicated by
\lstinline{use(}$m$\lstinline{,}$w'$\lstinline{,}$w-1$\lstinline{)} or
(ii)
also part of the time window~$w$, in which case $j_1 < j_2$ establishes
an asymmetric representation for the pair of operations in derived atoms
\lstinline{share((}$j_1$\lstinline{,}$s_1$\lstinline{),(}$j_2$\lstinline{,}$s_2$\lstinline{),}$p_1$\lstinline{,}$p_2$\lstinline{,}$x$\lstinline{,}$w$\lstinline{)}.
If the flag $x=\text{\lstinline{1}}$ signals that 
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)}
belongs to a previous time window~$w'$,
the rule in line~\ref{prg:base:ordered} % applies and 
derives the atom
\lstinline{order((}$j_1$\lstinline{,}$s_1$\lstinline{),(}$j_2$\lstinline{,}$s_2$\lstinline{),}$p_1$\lstinline{,}$w$\lstinline{)}
to signal that \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} needs to be completed
before performing \lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)}, i.e.,
the execution order must comply with the decomposition into time windows.
The rule in lines \ref{prg:base:same_job:begin}-\ref{prg:base:same_job:end} yields
a similar atom when \lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)}
is the predecessor operation $s_1 = s_2-1$ of \lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)}
in the same job $j_1=j_2$.
In contrast to the cases in which 
\lstinline{(}$j_1$\lstinline{,}$s_1$\lstinline{)} must be processed before
\lstinline{(}$j_2$\lstinline{,}$s_2$\lstinline{)},
the choice rule in line~\ref{prg:base:order}
allows for performing two operations sharing a machine in the lexicographic order
of their jobs if the operations belong to the same time window.
In case the atom representing execution in lexicographic order is not chosen,
the rule in lines \ref{prg:base:order:begin}-\ref{prg:base:order:end}
derives an atom signaling the inverse, given that the operations must not intersect and some sequence has to be determined.%
%
\lstinputlisting[float=t,label=prg:encoding,caption={Multi-shot ASP modulo DL encoding}]{listing/encoding.lp}

@@

%
Our first experiments, shown in Table~\ref{tab:Table01}, concern instances with $50\times15$, $50\times20$, and $100\times20$ jobs and machines with $750$, $1000$, or $2000$ operations, respectively. These operations were split into time windows with the M-EST decomposition strategy as a baseline because it leads to more robust solution quality than Job-based problem decompositions.
% KS removed, since it was already explained in Section 3.2
%The latter augments the encoding given in Listing~\ref{prg:decomposition} with bottleneck machines for inserting their operations into time windows, 
For each of the three problem sizes, we report the average makespan over ten instances, where smaller values indicate better schedules, the average runtime of \clingodl, and the average number of interrupted optimization runs on subproblems.
Each run was executed with a total timeout of $1000$ seconds, where each of $n$ subproblems got at most $1000/n$ seconds computation time.

For every set of instances, we gradually increase the number of time windows from~$1$ to~$6$ and additionally include results for~$10$ time windows to outline the trend of degrading solution quality when the partition into time windows becomes too fragmented for merging (optimized) partial schedules into proficient global solutions.
In fact, the shortest makespans, highlighted in boldface, were obtained with problem decomposition into $3$, $4$, or $6$ time windows, respectively, for the instances of size $50\times15$, $50\times20$, and $100\times20$.
Compared to the results for $1$ time window, 
which represent global optimization on the full problem,
we observe substantial makespan improvements, amounting to almost one order
of magnitude for the largest instances with $100\times20$ jobs and machines.
These advantages are not surprising, considering that the JSP instances are highly combinatorial \citep{shysha18a} and each global optimization run times out with a more or less optimized solution, where a good-quality schedule is the more challenging to find, the greater the number of operations gets.
Except for about 12\% completed optimization runs in multi-shot solving with $4$ time windows on the $50\times20$ instances, the best schedules were obtained with optimization runs timing out on subproblems. 
Nevertheless, they resulted in good-quality partial schedules that can be merged into better global solutions than achievable with more time windows, which become smaller and easier to solve but also cut off good-quality schedules.

\begin{table}[t]
  \caption{%
    Experiments varying the number of time windows on JSP benchmark sets of three sizes\label{tab:Table01}}%
    % Results for three sizes of job shop scheduling instances with different numbers of time windows using static machine-based EST decomposition \label{tab:Table01}}%
%    \mbox{~}\hfill
%    \resizebox{0.8\textwidth}{!}{%
%    \begin{minipage}{\textwidth}
  \centering
  \begin{tabular}{r l r r r r r r r}
  \topline%\hline
%       & {}        & \multicolumn{6}{c}{Group of Instances}  \\
%  &  &  \multicolumn{7}{c}{Number of Time Windows} \\ [2pt] \cline{3-9} 
  Size & Average & 1 & 2 & 3 & 4 & 5 & 6 & 10  % \\

@@

  {$50\times15$}       & Makespan             & $3542.1$   & $3149.7$	     &  \boldmath{$3083.8$} & $3110.7$ & $3213.6$ & $3225.3$ & $3524.9$ \\
  & Time             & $1000.0$	& $1000.0$   & $1000.0$	 & $934.4$ & $615.5$  & $243.8$  & $18.5$  \\
  & Interrupts    & $1.0$	    & $2.0$      & $3.0$	& $3.6$  & $2.7$  & $1.0$  & $0.1$       \\ [1.5mm]

@@

  % \\[-2.75mm]
  \botline%\hline
  \end{tabular}
 % }
%    \end{minipage}
% }
%    \hfill\hfill\mbox{~}
\end{table}

@@

\begin{table}[b]
    \caption{Experiments comparing Job- and Machine-based problem decomposition strategies\label{tab:Table02}}%
%    \mbox{~}\hfill
%    \resizebox{0.8\textwidth}{!}{%
%    \begin{minipage}{\textwidth}
    \centering
    \begin{tabular}{r r r r r r r r r r}
    \topline% \hline

@@

    % \\[-2.75mm]
    \botline% \hline
    \end{tabular}
   % }
%    \end{minipage}
% }
%    \hfill\hfill\mbox{~}
\end{table}
%
In Table~\ref{tab:Table02}, we provide a comparison of the Job- and Machine-based EST and MTWR approaches with %, providing the baseline abbreviated by M-EST.
%Further decomposition strategies are denoted by M-MTWR for the static Machine-based MTWR variant, 
dynamic versions of M-EST$^\text{\,d}$ and M-MTWR$^\text{\,d}$ that
determine M-EST and M-MTWR values w.r.t.\ partial schedules from previous time windows.
\ksverify{We excluded the dynamic variants of Job-based strategies, since they performed worse than their static originals.}
%, as well as static Job-based counterparts J-EST and J-MTWR in the first two rows.
For each decomposition strategy and JSP instance size, we present results for the number
of time windows,
which amounts to the next greater integer of the average number of interrupted optimization
runs in columns headed by Ints, leading to the shortest average makespan.
The considerably increased average makespans in the first two rows clearly indicate that
time windows determined with Job-based decomposition strategies are less adequate than
those investigating bottleneck machines in the first place, and then picking their operations
based on the smallest EST or greatest MTWR value as a secondary criterion.
While Machine-based decomposition turns out to be advantageous,
there is no clear gap between the static M-EST and M-MTWR variants as well as their dynamic
versions M-EST$^\text{\,d}$ and M-MTWR$^\text{\,d}$.
That is, the heuristic to schedule operations from highly loaded machines
as soon as possible works well for decomposing the considered JSP instances,
yet further solution quality differences due to minor deviations in the
decomposition strategy or optimization performance of \clingodl are negligible.
